---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First, we load the data from the working directory

```{r}
unzip("activity.zip")
data <- read.csv("activity.csv", stringsAsFactors=FALSE)

```

Then, we convert the dates into date format

```{r}
data$date <- as.Date(data$date)

```

The data then looks like this:

```{r}
head(data)
```


## What is mean total number of steps taken per day?

First we sum the data per day

```{r}
daydata <- aggregate(steps ~ date, data=data, FUN=sum)
```

Then generate a histogram of steps per day

```{r}
library(ggplot2)
pl1 <- qplot(daydata$steps, geom="histogram", xlab = "Steps per day",  binwidth = 1000) 
pl1
```

And calculate the mean and median of steps per day

```{r}
options(scipen=999, digits=2)
meanperday <- mean(daydata$steps, na.rm=TRUE)
medianperday <- median(daydata$steps, na.rm=TRUE)

```
The mean steps per day is `r meanperday`, and the median is `r medianperday`



## What is the average daily activity pattern?


First, we take a mean of the steps per interval (averaging across days)

```{r}
intervaldata <- aggregate(steps ~ interval, data=data, FUN=mean)
```

Then, make a time series plot of this data
```{r}
qplot(intervaldata$interval, intervaldata$steps, geom="line", xlab = "Interval", ylab= "Average steps taken in each 5-minute interval")
```

Then, find the interval with the maximum number of steps
```{r}
intervalmax <- intervaldata$interval[intervaldata$steps== max(intervaldata$steps)]
intervalmaxvalue <- max(intervaldata$steps)
```

The interval with the maximum average number of steps is `r intervalmax`, with an average of `r intervalmaxvalue` steps


## Imputing missing values

There are quite a few rows with missing values:
```{r}
length(data[is.na(data)==TRUE])
```

Let's fill all of the NA values in with the median for the interval.

I first create a lookup table for medians per interval

```{r}
lkp <- aggregate(steps ~ interval, data=data, FUN=median)
```

Then we create a new data frame called data2, with the interval and date columns, and for the steps column we:
1. first check if the data is NA using is.na
2. if it is, we take the value from the lkp data frame for that interval
3. if it isn't, then we take the original value (data$steps)


```{r}
data2 <- data.frame(interval = data$interval, date = data$date,
                    steps = ifelse (is.na(data$steps),
                                    (lkp$steps[lkp$interval == data$interval]),
                                    data$steps))
```

Then we take the same steps as we did on the original data, to aggregate, calculate the mean and median of steps per day, and generate a histogram of steps per day, that I show below the original data

```{r}
daydata2 <- aggregate(steps ~ date, data=data2, FUN=sum)
meanperday2 <- mean(daydata2$steps, na.rm=TRUE)
medianperday2 <- median(daydata2$steps, na.rm=TRUE)
pl2 <- qplot(daydata2$steps, geom="histogram", xlab = "Steps per day (corrected data)", binwidth = 1000) 
library(gridExtra)
grid.arrange(pl1, pl2, nrow=2)
```

The mean steps per day with this corrected data is `r meanperday2` (compared to `r meanperday`), and the median is `r medianperday2` (compared to `r medianperday2`).
So, the data does look slightly different, but not drastically.

This can be partially explained because the na data was only for 8 out of the 61 days. 

## Are there differences in activity patterns between weekdays and weekends?

First, I add in an additional column for whether the data is for a weekday or a weekend

```{r}
data$weekday <- "Weekday"
data$weekday[weekdays(data$date) %in% c("Saturday", "Sunday")] <- "Weekend"
data$weekday <- factor(data$weekday)
```

and then generate a time series plot

```{r}
intervaldata2 <- aggregate(steps ~ interval + weekday, data=data, FUN=mean)
plot <- ggplot(intervaldata2, aes(x=interval, y=steps, weekday)) + geom_line(color="darkblue") + facet_grid (weekday ~ .) + labs(title= "Average number of steps per interval: weekdays compared to weekends", x="5 minute interval")
plot
```